---
title: "DATA605_Final"
author: "Haig Bedros"
date: "2024-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### House Prices - Advanced Regression Techniques 

```{r}
library(readr)
```

```{r}
train_df <- read.csv('https://raw.githubusercontent.com/hbedros/House-Prices---Advanced-Regression-Techniques/main/data/train.csv')
train_df <- read.csv('https://raw.githubusercontent.com/hbedros/House-Prices---Advanced-Regression-Techniques/main/data/test.csv')
```

```{r}
head(train_df)
```

1. Pick one of the quanititative independent variables from the training data set (train.csv) , and define that variable as  X.   
2. Make sure this variable is skewed to the right!  
3. Pick the dependent variable and define it as  Y.

1. Quantitative Independent Variable (X)
Among your dataset fields, a good choice for a quantitative independent variable (predictor) is GrLivArea. This variable represents the above grade (ground) living area in square feet. This is a continuous variable and is likely to have a linear relationship with the property’s sale price, making it a suitable candidate for regression analysis. Let's define it as X:

But first, let's check if it's right skewed:
```{r}
library(e1071)

# Calculate skewness
skewness_grlivarea <- skewness(train_df$GrLivArea)

skewness_grlivarea
```

```{r}
# the skewness of GrLivArea is approximately 1.13, it already exhibits right skewness.

# defining X variable
X <- train_df$GrLivArea
```

2. Dependent Variable (Y)
The dependent variable, which you want to predict, is SalePrice. This is the property's sale price in dollars and is the target variable in your regression analysis. As you aim to predict the sale price based on other variables, it is appropriate to define it as Y:
```{r}
Y <- train_df$SalePrice
```

```{r}
# Scatter plot to examine the relationship
plot(X, Y, main="Scatter Plot", xlab="GrLivArea", ylab="SalePrice")

# Fit initial model
model <- lm(SalePrice ~ GrLivArea, data=train_df)
summary(model)

# Check residuals
plot(model$residuals, main="Residuals of Model", ylab="Residuals", xlab="Index")
hist(model$residuals, main="Histogram of Residuals", xlab="Residuals", breaks=30)

```

Probability.   Calculate as a minimum the below probabilities a through c.  Assume the small letter "x" is estimated as the 3d quartile of the X variable, and the small letter "y" is estimated as the 2d quartile of the Y variable.  Interpret the meaning of all probabilities.  In addition, make a table of counts as shown below.

a.  P(X>x | Y>y)		b.  P(X>x, Y>y)		c.  P(X<x | Y>y)		

```{r}
# X <- train_df$GrLivArea
# Y <- train_df$SalePrice

# Load your data
# train_df <- read.csv("path/to/your/data.csv")

# Step 1: Calculate quartiles
x_3rd_quartile <- quantile(X, 0.75)
y_2nd_quartile <- quantile(Y, 0.5)

# Step 2: Create flags
train_df$X_gt_x <- ifelse(X > x_3rd_quartile, 1, 0)
train_df$Y_gt_y <- ifelse(Y > y_2nd_quartile, 1, 0)

# Step 3: Build contingency table
table <- table(train_df$X_gt_x, train_df$Y_gt_y)
print(table)

# Step 4: Calculate probabilities
# P(X > x | Y > y)
p_X_gt_x_given_Y_gt_y <- table[2, 2] / sum(table[, 2])

# P(X > x, Y > y)
p_X_gt_x_and_Y_gt_y <- table[2, 2] / sum(table)

# P(X < x | Y > y)
p_X_lt_x_given_Y_gt_y <- table[1, 2] / sum(table[, 2])

# Print probabilities
# print(paste("x_3rd_quartile:", x_3rd_quartile))
# print(paste("y_2nd_quartile:", y_2nd_quartile))
print(paste("P(X > x | Y > y):", p_X_gt_x_given_Y_gt_y))
print(paste("P(X > x, Y > y):", p_X_gt_x_and_Y_gt_y))
print(paste("P(X < x | Y > y):", p_X_lt_x_given_Y_gt_y))

```

```{r}
# Assuming train_df is your correct data frame
library(dplyr)

# Calculate quartiles
x_3rd_quartile <- quantile(train_df$GrLivArea, 0.75, na.rm = TRUE)
y_2nd_quartile <- quantile(train_df$SalePrice, 0.5, na.rm = TRUE)

# Add classifications based on quartiles
train_df <- train_df %>%
  mutate(
    X_class = if_else(GrLivArea <= x_3rd_quartile, "<=3d quartile", ">3d quartile"),
    Y_class = if_else(SalePrice <= y_2nd_quartile, "<=2d quartile", ">2d quartile")
  )

# Build contingency table
contingency_table <- table(train_df$X_class, train_df$Y_class)

# Adding row and column totals
contingency_table <- addmargins(contingency_table)

# Display the contingency table
print(contingency_table)

```